---
title: Bio4158/Bio5158 - Report 4
subtitle: Exercise
author: "Group 22 - Nalan Kurnaz"
date: today
format:
  html:
    embed-resources: true
    toc: true
    code-fold: show
    code-tools: true
---


# The problem

The peppered moth (_Biston betularia_ Lepidoptera : Geometridae) is a nocturnal insect common in Europe and North America. Adults fly at night and stay put during the day on birch trees. Adult coloration is variable, ranging from a clear form (typica) to a melanic form that is very dark (carbonaria), with many intermediate forms.

The proportion of individuals of each form has varied considerably in natural populations over the last 150 years. The melanic form was rare before 1850 but represented over 90% of the population in industrial sectors of England by 1900. According to Majerus (1998), J.W. Tutt was the first to suggest in 1896 an hypothesis to explain the temporal change in the proportion of the dark form. 

__Biological hypothesis__: Differences in the relative frequency of the two forms arise because survival from bird predation of each depends on habitat color. In particular, the pale form is less visible in non-polluted environments when birch trees are covered by pale greyish lichens. Inversely, the dark form is favored in industrial areas where soot darkens their resting places. Therefore, the increase of the dark form (carbonaria) in England resulted from bird predation and a change in habitat (soot pollution). 

# The study

In a study designed to test this hypothesis, a random sample of 30 insectivorous birds was used to estimate predation susceptibility of the pale and dark forms on both pale and dark substrates. **Two growth chamber** were set up containing branches and leaves:

1. **in the first, the branches and leaves were taken from a polluted area and were covered with soot (and were thus dark)**;
2. **in the second, the branches and leaves were taken from a non-polluted area and were thus lighter in colour**.

Each bird was introduced in turn into each chamber (**half into the dark first** and **half into the light first**), with a fasting period in between to keep hunger constant. Just before introducing a bird to a chamber, **50 pale moths** (typica) and **50 dark moths** (carbonaria) were randomly selected and put in the chamber. After a fixed period of time, the number of pale and dark moths captured by the bird in the pale and dark chambers was determined. After each experiment, the remaining moths were discarded before putting in fresh moths for the next experiment.


# The data

The file `Biston_student.csv` contains a fraction of the data that a professor has managed to unearth after his MSc student left without a trace prior to publishing their results. Unfortunately, the data sheets are **incomplete** and do not allow him to tell which value comes from which bird. In this file **there are 30 lines** (representing the 30 birds) and **two columns** (**one for the number of dark moths captured in the dark chamber and one for the number of dark moths captured in the pale chamber*). However, **numbers on a given line do NOT necessarily come from the same bird (i.e. the order of observations in the two columns has been mixed up)**.

The file `Biston_pd2.csv` is later reconstructed by a post-doc from the same lab. This post-doc is much more persistent than the prof and manages to find additional data sheets that allow her to match each result to a particular bird. This file has the same structure as `Biston_student.csv`, and contains the same data, **except that observations on the same line now come from the same bird (i.e. paired observations are known)**. 


# The task
### Data Preprocessing
```{r}
#import the libraries and read the data
#install.packages("car")
library(dplyr)
library(pwr)
library(ggplot2)
library(lmtest)
library(car)
library(boot)


student_data <- read.csv("https://raw.githubusercontent.com/nalank1/r-files/refs/heads/main/data-bioxx58/Biston_student.csv", sep = ",", stringsAsFactors = TRUE)

postdoc_data <- read.csv("https://raw.githubusercontent.com/nalank1/r-files/refs/heads/main/data-bioxx58/Biston_pd_2.csv", sep = ",", stringsAsFactors = TRUE)

```

## 1. Student data

Analyze the data and determine if the results in `Biston_student.csv` support or refute the biological hypothesis. 


### a.

::: {.callout}
Give a statement of the biological hypothesis and its prediction (1 point)
:::

<!--#################################################################-->

The survival rate of dark peppered moths depends on the color of the substrate moth is located in. Dark moths are expected to be less frequently captured by birds in dark (polluted) environment where they are better camouflaged and more frequently captured in pale (non-polluted) environments, where they are more visible. It was predicted that fewer dark moths will be captured in the dark chamber compared to the pale chamber.

<!--#################################################################-->

### b.

::: {.callout}
a statement of the null hypothesis (1 point)
:::

<!--#################################################################-->

The mean number of dark peppered moths captured in dark chamber is equal the mean number of dark peppered moths captured in pale chamber. 

<!--#################################################################-->

### c.

::: {.callout}
fit the adequate statistical model, and evaluate the statistical assumptions _(2 points)_ - not paired, one tailed test
:::

<!--#################################################################-->
```{r}
#do not change into ordinal - you will lose the clarity
ggplot(student_data, aes(x = room, y = number, fill = room)) +
  geom_boxplot() +
  labs(title = "Number of Dark Peppered Moths Captured by Chamber",
       x = "Chamber",
       y = "Number of Dark Moths Captured") +
  theme_minimal()
```

```{r}
#checking normality
ggplot(student_data, aes(x = number, fill = room)) +
  geom_histogram(aes(y = ..density..), color = "black", alpha = 0.3, bins = 10) +
  geom_density(alpha = 0.5) +
  geom_rug() +
  stat_function(
    fun = dnorm,
    args = list(
      mean = mean(student_data$number),
      sd = sd(student_data$number)
    ),
    color = "red"
  ) +
  facet_grid(. ~ room) + #splits the plot into separate panels for each chamber
  xlab("Number of Dark Moths Captured") +
  theme_minimal()
```
```{r}
shapiro.test(student_data$number)
```

As seen in both histograms and Shapiro-Wilk tests, it was indicated that both groups are approximately normally distributed (p-value > 0.05) so parametric t-test is appropriate for comparing means.

As groups are normally distributed and the variances in the two groups don't differ much, it is safe to use two sample t-test.

```{r}
#assuming variances are equal
t_result <- t.test(number ~ room, data = student_data, alternative = "two.sided", var.equal = TRUE)
t_result
t_critical <- qt(p = 0.975, df = 58)
t_critical
```

The t-test showed a non-significant difference between groups (p = 0.126). Although the sample means differ (29.47 vs 32.27), this difference is not statistically significant and could easily occur by chance alone. We want to also be sure and test all bases. 

```{r}
#assuming variances are not equal
t.test(number ~ room, data = student_data, alternative = "two.sided", var.equal = FALSE)
```


```{r}
wilcox.test(number ~ room, data = student_data, alternative = "two.sided")
```

The Welch Two Sample t-test produced results nearly identical to the standard t-test, with only minor differences in the confidence interval and degrees of freedom. Both t-tests yield a t-statistic of -1.55, which falls below the critical value of approximately 2.00 required for statistical significance at α = 0.05. This indicates that the observed difference in means is not large enough to be considered statistically significant.

The non-parametric Wilcoxon rank sum test aligns with this conclusion, also failing to find sufficient evidence for a difference in distributions between the two groups.

Given that these different approaches all point toward the same conclusion, a final assessment of the model assumptions was conducted to verify the appropriateness of the parametric t-test.

```{r}
model.1 <- lm(number ~ room, data = student_data)
par(mfrow = c(2,2))
plot(model.1)

```

The first graph shows the spread of the residuals for the two groups and since both columns have roughly the same vertical height (similar range of residuals), the variances are homogeneous. The second graph (QQ plot of the residuals of the model) also seems to depict a straight line, meaning the normality assumption is satisfied. The proof that normality assumption is satisfied can also be verified by the Shapiro Test on residuals below. With W = 0.97992 and p-value = 0.4251 (>0.05), the test indicates that the residuals are normal. The other two tests confirm the assessment of homogeneity of variance (homoscedasticity). The Levene's Test gives a p-value of 0.711. The Breusch-Pagan test (bptest) gives a p-value of 0.8673. Since both p-values (0.711 and 0.8673) are much greater than 0.05, we fail to reject the null hypothesis of equal variances. This statistically confirms that the homogeneity of variance assumption is met. 

```{r}
shapiro.test(residuals(model.1))

leveneTest(model.1)

bptest(model.1)
```


<!--#################################################################-->

### d.
::: {.callout} 
The statistical conclusion (i.e. inference). Do you accept or reject the null hypothesis? (3 points)
:::

<!--#################################################################-->

Based on the results of the two-sample t-test, we fail to reject the null hypothesis that the mean number of dark peppered moths is the same in dark and pale environments. The decision was made by comparing the calculated t-statistic to the critical t-value from the Student's t-distribution. The calculated t-statistic from the test is |t_statistic| = 1.5532, and for a two-tailed test with a significance level of \alpha\ = 0.05 and df = 58, the t_critical is 2.002. Since |t_statistic| < t_critical, we fail to reject the null hypothesis. <br>
The validity of results of t-test were verified by the model's underlying assumptions. Shapiro-Wilk test was used on the data and the residuals,and QQ Plot indicated no significant deviation from normality. Levene's Test (F = 0.139 and p = 0.711) and BP Test (BP = 0.028, p = 0.867) confirm the assumption of equal variances. The residual plot also showed the consistent spread for both groups. <br>

Additionally, we have also tried additional t-test under unequal variances and non-parametric Wilcoxon test. Both of these tests failed to reject null hypothesis, supporting the conclusion from the primary analysis.

<!--#################################################################-->

### e.
::: {.callout} 
The “biological” conclusion (i.e. a qualified statement about the biological hypothesis concerning moth survival, including an estimate of raw effect size, its precision, and whether an effect of this magnitude seems biologically interesting) (2 points).
:::

<!--#################################################################-->
Based on the student_data analysis, we fail to find statistical support for the biological hypothesis that substrate color affects the survival of dark peppered moths through bird predation. The raw effect size (the observed difference in mean moth counts) was 2.8 moths with the pale room showing a slightly higher mean (32.27) than the dark room (29.47). However, this small observed effect is highly imprecise. The 95% confidence interval for the true difference ranges from -6.41 to +0.81 moths, meaning the data are consistent with a scenario where the dark room could have up to 6.4 fewer moths, both rooms could be equal or dark room could have a slight advantage of 0.8 moths. Given that the confidence interval includes zero and spans both positive and negative values, an effect of this magnitude is not statistically distinguishable from random variation. Although a consistent 2.8-moth difference (approximately a 9% change from the mean) would likely be biologically meaningful as a selective pressure, the imprecision of this estimate means we cannot conclude such an effect exists. Therefore, this analysis fails to provide conclusive evidence to support the hypothesis.

<!--#################################################################-->

### f.
::: {.callout} 
A statement about your confidence in the statistical conclusion (for example with respect to the violation of certain assumptions, to potential biases created by missing data, to suboptimal experimental design, and/or to power issues ) (2 points)
:::

<!--#################################################################-->
We have low confidence in the statistical conclusion to fail to reject the null hypothesis as it is supported by multiple consistent analyses but limited by the study’s design and low power. First, all statistical assumptions for the t-test were thoroughly validated; both graphical analysis and formal tests confirmed no significant violations of normality or homoscedasticity. Second, the conclusion was consistently reinforced by alternative approaches, including Welch's t-test and the non-parametric Wilcoxon test. The main limitation is the issue of low statistical power. A post-hoc power analysis reveals the study had only 33.3% power to detect the observed effect size (d = 0.40) with the given sample size (n=30 per group). This means that even if a true difference exists, there was a high probability of not detecting it (Type II error). Therefore, while the analyses consistently support the null hypothesis, our confidence in interpreting this as evidence of no biological effect is low. <br>

Bootstrap analysis with 1,000 replicates produced a 95% confidence interval of (-0.619, 6.013) for the difference in moth captures between pale and dark rooms. Since this interval includes zero, it confirms that the observed difference of 2.8 moths is not statistically significant. As this interval includes zero and spans both positive and negative values, it supports the statistical conclusion but also highlights the uncertainty due to limited sample size and the unpaired design. The statistical conclusion may be consistent, however the external confidence is reduced by the limited power and unpaired data structure.

```{r}
mean_dark <- t_result$estimate[1]
mean_pale <- t_result$estimate[2]

sd_dark <- sd(student_data$number[student_data$room == "dark"])
sd_pale <- sd(student_data$number[student_data$room == "pale"])

d <- function(u1, u2, sd1, sd2){
  abs(u1 - u2) / sqrt((sd1^2 + sd2^2) / 2)
}

pwr.t.test(n = 30, d = d(u1 = mean_dark, u2 = mean_pale, sd1 = sd_dark, sd2 = sd_pale), sig.level = 0.05, type = "two.sample")
```

```{r}
bs <- function(formula, data, indices){
  d <- data[indices, ]
  fit <- lm(formula, data = d)
  return(coef(fit))
}

results <- boot(
  data = student_data, statistic = bs, R = 1000,
  formula = number ~ room
)
results
boot.ci(results, type = "bca", index = 2)
```

<!--#################################################################-->

## 2. Postdoc data

Analyze the data and determine if the results in `Biston_pd2.csv `support or refute the biological hypothesis.

### a.
::: {.callout} 
a statement of the biological hypothesis and its prediction (1 point)
:::

<!--#################################################################-->

The mean number of dark peppered moths in pale and dark environment differs because the survival from bird predation depends on the colour of the moth in relation to the colour of the environment. Dark peppered moths are better camouflaged on dark substrates than on pale (non‑polluted) substrates, so a bird will capture fewer dark moths when foraging on a dark substrate than when foraging on a pale substrate. <br>

When the same individual birds are tested in both environments, they will capture significantly fewer dark moths in the dark, polluted environment (where the dark moths are less visible) than in the pale, non-polluted environment.

<!--#################################################################-->

### b.
::: {.callout} 
a statement of the null hypothesis (1 point)
:::

<!--#################################################################-->
The null hypothesis is that the mean difference in the number of dark peppered moths captured between the pale and dark chambers is zero. 

<!--#################################################################-->

### c.
::: {.callout} 
fit the adequate statistical model, and evaluate the statistical assumptions _(3 points)_
:::

<!--#################################################################-->
```{r}
postdoc_data$difference <- postdoc_data$pale - postdoc_data$dark

par(mfrow = c(1, 1))

density_values <- density(postdoc_data$difference)

max_density <- max(density_values$y)

hist(postdoc_data$difference,
     main = "Histogram of Differences",
     xlab = "Difference (Pale - Dark)",
     col = "lightblue",
     probability = TRUE,
     ylim = c(0, max_density * 1.1)) # TRUE helped me to overlay the density curve and normal curve on same scale

lines(density_values, col = "red", lwd = 2)
curve(dnorm(x, 
            mean = mean(postdoc_data$difference),
            sd = sd(postdoc_data$difference)),
      col = "blue", lwd = 2, lty = 2, add = TRUE)
legend("topright", legend = c("Data Density", "Normal Curve"),
      col = c("red", "blue"), lty = c(1, 2), lwd = 2)
```

```{r}
qqnorm(postdoc_data$difference, main = "QQ-Plot of Differences")
qqline(postdoc_data$difference, col = "red")
```

```{r}
boxplot(postdoc_data$difference, 
        main = "Boxplot of DIfferences",
        ylab = "Difference (Pale - Dark)",
        col = "lightgreen")
points(mean(postdoc_data$difference), pch = 20, col = "red", cex = 1.5)
legend("topright", legend = "Mean", pch = 20, col = "red")
```

```{r}

plot(postdoc_data$difference ~ rep(mean(postdoc_data$difference), nrow(postdoc_data)),
    main = "Differences vs Central Tendency",
    xlab = "Overall Mean Difference",
    ylab = "Individual Differences",
    pch = 20, col = "darkblue")

abline(h = 0, col = "red", lwd = 2)
abline(h = mean(postdoc_data$difference), col = "blue", lwd = 2, lty = 2)
legend("topright", 
      legend = c("Zero Reference", "Mean Difference"),
      col = c("red", "blue"),
      lty = c(1, 2),
      lwd = 2)

```

```{r}
shapiro.test(postdoc_data$difference)
```

A one-tailed paired t-test was selected as the appropriate statistical model because the experiment had repeated measures, meaning each bird is tested in both chambers. Also, we were testing a directional biological prediction since more dark moths captured in pale chambers. <br>

The statistical assumptions were evaluated, beginning with the normality of the differences between paired observations. This was assessed using multiple complementary approaches: a histogram with overlaid density curves showed a bell-shaped distribution that closely followed the theoretical normal curve; a Q-Q plot showed points falling reasonably close to the reference line; a boxplot displayed a symmetric distribution with no evidence of extreme outliers and a formal Shapiro-Wilk test yielded a p-value of 0.1215, above the 0.05 threshold, confirming no significant departure from normality. The independence assumption was satisfied as well since birds were randomly sampled and tested independently. Both the normality and independence assumptions are met, validating the use of the paired t-test for this analysis.

```{r}
t.test(postdoc_data$pale, postdoc_data$dark, paired = TRUE, alternative = "greater")
```


<!--#################################################################-->

### d.
::: {.callout} 
The statistical conclusion (i.e. inference). Do you accept or reject the null hypothesis? Include your test results. (3 points)
:::

<!--#################################################################-->
Based on the results of the paired t-test, we reject the null hypothesis that the mean difference in dark moth captures between chambers is zero. The paired t-test yielded a statistically significant result (t_statistic(29) = 2.0932, p-value = 0.0226). Also, the 95% lower confidence bound (0.527 moths) excludes zero, indicating the true effect is positive. In other words, having confidence interval above zero means the effect is consistently positive and zero is not a value for the true effect size. Since |t_statistic| = 2.0932 > |t_critical| = qt(0.95, df = 29) = 1.699, we are safe to reject the null hypothesis. 

<!--#################################################################-->

### e.
::: {.callout} 
The “biological” conclusion (i.e. a qualified statement about the biological hypothesis concerning moth survival, including an estimate of raw effect size, its precision, and whether an effect of this magnitude seems biologically interesting) (2 points).
:::

<!--#################################################################-->

The results supports the biological hypothesis that the survival of dark peppered moths (carbonaria) is significantly influenced by background color due to bird predation. The experiment showed a statistically significant advantage for the dark moths when resting on the dark, polluted background. The raw effect size is a mean difference of 2.8 dark moths captured per bird. In other words, on average, birds captured 2.8 more dark moths in the pale (non-polluted) chamber (mean capture 32.27) than in the dark (polluted) chamber (mean capture 29.47). This observed direction is precisely what is predicted by the camouflage hypothesis. The effect is estimated with reasonable precision, as evidenced by the paired t-test's p-value (0.0226) and the 95% lower confidence bound of 0.53 moths. Because the entire confidence range for the one-tailed test is [0.53,inf], we can be confident that the true difference is a positive value, ruling out a zero or negative difference. From a biological perspective, an average difference of 2.8 moths out of 30 available represents a 9.3% change in survival rate in a single foraging event. This level of selective pressure can be considered biologically strong for the rapid evolutionary change observed in the peppered moth population. The paired design was critical here successfully controlling for bird-to-bird variability to isolate and detect the effect.


<!--#################################################################-->

### f.
::: {.callout} 
A statement about your confidence in the statistical conclusion (for example with respect to the violation of certain assumptions, to potential biases created by missing data, to suboptimal experimental design, and/or to power issues ) (2 points)
:::

<!--#################################################################-->

We have strong confidence in our conclusion that birds catch more dark moths in pale chambers, leading us to confidently reject the null hypothesis. Our confidence stems from the paired experiments, which effectively controlled for individual bird variability, and from meeting all key statistical assumptions including normality of differences and data independence. A post-hoc power analysis indicated that the study had 65.5% power to detect the observed medium effect size (d = 0.38). The analysis yielded a statistically significant result (p=0.0226). Overall, the paired design allowed the true biological difference to be detected. 

```{r}
sd_postdoc <- sd(postdoc_data$difference)
cohen_d <- mean(postdoc_data$difference) / sd_postdoc

power_res <- pwr.t.test(
  n = 30,
  d = cohen_d,
  sig.level = 0.05,
  type = "one.sample",
  alternative = "greater"
)
power_res
```

<!--#################################################################-->

## 3. Power issues

Assuming that a difference of 5% (or 2.5 moths) of predation rate between treatments is biologically meaningful, what would be the power to detect this with alpha=0.05 and the data collected by:

### a.
::: {.callout} 
the student? _(1 point)_
:::

<!--#################################################################-->
```{r}
d_students <- 2.5 / sqrt((sd_dark^2 + sd_pale^2) / 2)

power_students <- pwr.t.test(
  n = 30,
  d = d_students,
  sig.level = 0.05,
  type = "two.sample",
  alternative = "two.sided"
)

power_students
```

For the student's unpaired design with 30 birds in each group, the power to detect a biologically meaningful difference of 2.5 moths is approximately 27.6%.

<!--#################################################################-->

### b.
::: {.callout} 
the postdoc? _(1 point)_
:::

::: callout-warning
`d` for a paired t-test is estimated as $d = \text{mean differences} / \text{sd differences}$

using the `d()` function you can specify it as `d(0, mean differences, sd differences, sd differences)`
:::

<!--#################################################################-->
```{r}
d_postdoc <- d(u1 = 2.5, u2 = 0, sd1 = sd(postdoc_data$difference), sd2 = sd(postdoc_data$difference))

power_postdoc <- pwr.t.test(
  n = 30,
  d = d_postdoc,
  sig.level = 0.05,
  type = "paired",
  alternative = "greater"
)

power_postdoc
```

For the postdoc's paired design with 30 bird pairs, the power to detect the same biologically meaningful difference of 2.5 moths is approximately 57.1%

<!--#################################################################-->

### c.
::: {.callout} 
Comment the difference between the power using the student and postdoc data. _(1 point)_
:::

<!--#################################################################-->

The paired postdoc design (57.1% power) provides substantially higher detection capability than the unpaired student design (27.6% power). This twice of increase in power occurs because the paired design reduces variability by controlling for individual bird differences, yielding a more sensitive test. 

<!--#################################################################-->
