---
title: Bio4158/Bio5158 - Report 2
subtitle: Exercise
author: "Group 22 - Nalan Kurnaz"
date: today
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    embed-resources: true
    code-fold: show
    code-tools: true
---

## The problem 

Black fly larvae (Diptera: Simuliidae) were sampled in February at the
outlet of two lakes of the Eastern Townships (Lakes Orford and
Lovering). The body length of each larva was measured. The data are in
the file `simulies.csv` The relationship between body length (L, in mm)
and dry mass (M, in $\mu$g) for the dominant species (*P.
mixtum/fuscum*) is:

$$
M = 1.36 L^{3.05}
$$

Analyze the data and answer the questions. *Total = 20 points*

The effect size `d` is estimate as:

$$
d = \frac{\left|\mu_2 - \mu_1\right|}{\sqrt{\frac{sd_1^2 + sd_2^2}{2}}}
$$

## The task

### Site specific summary statistics

#### Data Preprocessing
```{r}
#importing libraries, loading the data, reading it, data wrangling
install.packages("pwr")
library(dplyr)
library(pwr)

simulies_data <- read.csv("data-bioxx58/simulies.csv", header=TRUE, sep=",", stringsAsFactors=TRUE)
print(simulies_data)
str(simulies_data)
sum(is.na(simulies_data))

```
<!--#################################################################-->


::: {.callout}
# To do
Calculate the mean mass and its standard deviation at each site. *(4 points)*
:::

<!--#################################################################-->
```{r}
for (i in 1:length(simulies_data$Length)){
  simulies_data$Mass[i] <- 1.36* (simulies_data$Length[i]**(3.05))

}
summary <- simulies_data %>%
group_by(Site) %>%
summarize(mean_mass = mean(Mass), standard_deviation = sd(Mass))

summary

```

The mean of Lovering and Orford are 1.73 and 2.60, respectively. The standard deviation for Lovering and Orford are 0.795 and 1.22, respectively.

<!--#################################################################-->



### Power 

::: {.callout}
# To do
Using the standard deviations observed at each site, calculate the power of a two-tailed t test for independent means *(4 points)*
:::

####
::: {.callout}
# To do
if the mass difference was 5 $\mu$g, $\alpha$=0.05, and that 100 larvae were sampled at each site
:::

<!--#################################################################-->
```{r}
#write your code here
## calculating the power of a two-tailed t test

orford_mean <- summary %>% 
filter(Site == "Orford") %>% pull(mean_mass)
orford_sd <- summary %>%
filter(Site == "Orford") %>% pull(standard_deviation)

lovering_mean <- summary %>%
filter(Site == "Lovering") %>% pull(mean_mass)
lovering_sd <- summary %>%
filter(Site == "Lovering") %>% pull(standard_deviation)

# d <- abs(orford_mean - lovering_mean) / sqrt(((orford_sd**2) + (lovering_sd**2))/2)

mass_difference <- 5 / sqrt(((orford_sd**2) + (lovering_sd**2))/2)

pwr <- pwr.t.test(n = 100, mass_difference, sig.level = 0.05, type = "two.sample", alternative = "two.sided")
print(pwr)


```

Since Power is 1 and Power = 1-\beta\, we found \beta\ to be 0. \beta\ means the probability of making Type II error, in other words, the probability of failing to reject the null hypothesis when alternative hypothesis is true is zero. Therefore, it can be verified that we can reject null hypothesis. 

<!--#################################################################-->

####

::: {.callout}
# To do
if the mass difference was 20 $\mu$g, $\alpha$=0.05, and that 100 larvae were sampled at each site
:::

<!--#################################################################-->
```{r}
#write your code here

mass_difference <- 20 / sqrt(((orford_sd**2) + (lovering_sd**2))/2)
pwr2 <- pwr.t.test(n = 100, mass_difference, sig.level = 0.05, type = "two.sample", alternative = "two.sided")
pwr2
```

Write your explanation

<!--#################################################################-->


####

::: {.callout}
# To do
if the mass difference was 50 $\mu$g, $\alpha$=0.05, and that 100 larvae were sampled at each site
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->



####

::: {.callout}
# To do
Comment on how changing the mass difference affects power.
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->



### Required sample size 
::: {.callout}
# To do
Calculate the required sample size to detect, with a two-tailed t-test for independent means, a difference of 50 $\mu$g between the means *(4 points)*
:::

####

::: {.callout}
# To do
with a power of 80% and $\alpha$=0.05
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->


####

::: {.callout}
# To do
with a power of 80% and $\alpha$=0.001
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->


####
::: {.callout}
# To do
with a power of 95% and $\alpha$=0.
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->



####

::: {.callout}
# To do
Comment on how changing power and $\alpha$ affects the required sample size.
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->



### Detectable effect size

::: {.callout}
# To do
Calculate the detectable effect size (`d`) with a two-tailed t-test on independent means, given the observed standard deviations *(4 points)*
:::

####
::: {.callout}
# To do
With a power of 80%, $\alpha$=0.05, and measurements on 10 larvae at each site
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->


####
::: {.callout}
# To do
With a power of 80%, $\alpha$=0.05, and measurements on 200 larvae at each site
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->


####
::: {.callout}
# To do
With a power of 80%, $\alpha$=0.05, and measurements on 20 larvae at one site and 380 larvae at the second site
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->

####
::: {.callout}
# To do
Comment on how changing sample sizes affect `d`.
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->


### Differences in average mass

::: {.callout}
# To do
Calculate the difference in average mass, in $\mu$g, that can be detected given your estimate of the minimum detectable effect size (`d`) at 2.4.1-2.4.3 *(4 points)*
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->

