---
title: Bio4158/Bio5158 - Report 2
subtitle: Exercise
author: "Group 22 - Nalan Kurnaz"
date: today
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    embed-resources: true
    code-fold: show
    code-tools: true
---

## The problem 

Black fly larvae (Diptera: Simuliidae) were sampled in February at the
outlet of two lakes of the Eastern Townships (Lakes Orford and
Lovering). The body length of each larva was measured. The data are in
the file `simulies.csv` The relationship between body length (L, in mm)
and dry mass (M, in $\mu$g) for the dominant species (*P.
mixtum/fuscum*) is:

$$
M = 1.36 L^{3.05}
$$

Analyze the data and answer the questions. *Total = 20 points*

The effect size `d` is estimate as:

$$
d = \frac{\left|\mu_2 - \mu_1\right|}{\sqrt{\frac{sd_1^2 + sd_2^2}{2}}}
$$

## The task

### Site specific summary statistics

#### Data Preprocessing
```{r}
#importing libraries, loading the data, reading it, data wrangling
#install.packages("pwr")
library(dplyr)
library(pwr)
library(ggplot2)

simulies_data <- read.csv("data-bioxx58/simulies.csv", header=TRUE, sep=",", stringsAsFactors=TRUE)
str(simulies_data)
sum(is.na(simulies_data))

```
<!--#################################################################-->


::: {.callout}
# To do
Calculate the mean mass and its standard deviation at each site. *(4 points)*
:::

<!--#################################################################-->
```{r}
for (i in 1:length(simulies_data$Length)){
  simulies_data$Mass[i] <- 1.36* (simulies_data$Length[i]**(3.05))

}
summary <- simulies_data %>%
group_by(Site) %>%
summarize(mean_mass = mean(Mass), standard_deviation = sd(Mass))

summary

```

The mean mass of Lovering and Orford is 13.1 and 43.0, respectively. Their standard deviation is 23.8 and 48.0 for Lovering and Orford, respectively. 'WRITE MORE EXPLANATION'

<!--#################################################################-->



### Power 

::: {.callout}
# To do
Using the standard deviations observed at each site, calculate the power of a two-tailed t test for independent means *(4 points)*
:::

####
::: {.callout}
# To do
if the mass difference was 5 $\mu$g, $\alpha$=0.05, and that 100 larvae were sampled at each site
:::

<!--#################################################################-->
```{r}

## calculating the power of a two-tailed t test

orford_mean <- summary %>% 
filter(Site == "Orford") %>% pull(mean_mass)
orford_sd <- summary %>%
filter(Site == "Orford") %>% pull(standard_deviation)

lovering_mean <- summary %>%
filter(Site == "Lovering") %>% pull(mean_mass)
lovering_sd <- summary %>%
filter(Site == "Lovering") %>% pull(standard_deviation)

# d <- abs(orford_mean - lovering_mean) / sqrt(((orford_sd**2) + (lovering_sd**2))/2)

d1 <- 5 / sqrt(((orford_sd**2) + (lovering_sd**2))/2)

pwr <- pwr.t.test(n = 100, d1, sig.level = 0.05, type = "two.sample", alternative = "two.sided")
print(pwr)


```

Based on the sample size = 100, effect size = 0.1319744 and alpha = 0.05, the power of the two-tailed t-test is 0.1531382.

<!--#################################################################-->

####

::: {.callout}
# To do
if the mass difference was 20 $\mu$g, $\alpha$=0.05, and that 100 larvae were sampled at each site
:::

<!--#################################################################-->
```{r}

d2 <- 20 / sqrt(((orford_sd**2) + (lovering_sd**2))/2)
pwr2 <- pwr.t.test(n = 100, d2, sig.level = 0.05, type = "two.sample", alternative = "two.sided")
pwr2
```

Based on the sample size = 100, effect size = 0.5278978 and the significance level = 0.05, the power of the two-tailed t-test is 0.9603422.

<!--#################################################################-->


####

::: {.callout}
# To do
if the mass difference was 50 $\mu$g, $\alpha$=0.05, and that 100 larvae were sampled at each site
:::

<!--#################################################################-->
```{r}
d3 <- 50 / sqrt(((orford_sd**2) + (lovering_sd**2))/2)

pwr3 <- pwr.t.test(n = 100, d3, sig.level=0.05, type = "two.sample", alternative = "two.sided")

pwr3
```

Based on the sample size = 100, effect size = 1.319744 and the significance level = 0.05, the power of the two-tailed t-test is 1.
<!--#################################################################-->



####

::: {.callout}
# To do
Comment on how changing the mass difference affects power. - make sure to show this via plot
:::

<!--#################################################################-->
```{r}
#write your code here
# d1 <- mass_difference
# x <- seq(-4, 4, length=200)
# y0 <- dt(x, 198)
# y1 <- dt(x, 34, ncp = d1 * sqrt(200) / 2)
# plot(x, y0, type="l", col = "red", lwd = 2) #H0 distribution
# qc <- qt(0.025, 198)
# abline(v = qc, col = "green")
# abline(v = -qc, col = "green")
# lines(x, y1, type="l", col = "blue", lwd = 2)

# polygon(
#   c(x[x <= -qc], -qc),
#   c(y1[x <= -qc], 0),
#   col = rgb(red = 0, green = 0.2, blue = 1, alpha = 0.5)
# )

plot_power <- function(d_value, title = "Power Plot"){
  df <- 198
  x <- seq(-4, 8, length = 200)
  y0 <- dt(x, df) #H0
  y1 <- suppressWarnings(dt(x, df, ncp = d_value * sqrt(200)/2)) #H1
  plot(x, y0, type = "l", col = "red", lwd = 2, ylim = c(0, max(y0, y1) * 1.1), main = title, ylab = "Density", xlab = "t-value")
  qc <- qt(0.025, df)
  abline(v = qc, col = "green")
  abline(v = -qc, col = "green")

  lines(x, y1, type = "l", col = "blue", lwd = 2, lty = 2)

  polygon(
    c(x[x <= -qc], -qc),
    c(y1[x <= -qc], 0),
    col = rgb(0, 0.2, 1, alpha=0.5)
  )
}

par(mfrow = c(3,1))
plot_power(d1, title = "Mass Difference = 5")
plot_power(d2, title = "Mass Difference = 20")
plot_power(d3, title = "Mass Difference = 50")
```

Write your explanation

<!--#################################################################-->



### Required sample size 
::: {.callout}
# To do
Calculate the required sample size to detect, with a two-tailed t-test for independent means, a difference of 50 $\mu$g between the means *(4 points)*
:::

####

::: {.callout}
# To do
with a power of 80% and $\alpha$=0.05
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->


####

::: {.callout}
# To do
with a power of 80% and $\alpha$=0.001
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->


####
::: {.callout}
# To do
with a power of 95% and $\alpha$=0.
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->



####

::: {.callout}
# To do
Comment on how changing power and $\alpha$ affects the required sample size.
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->



### Detectable effect size

::: {.callout}
# To do
Calculate the detectable effect size (`d`) with a two-tailed t-test on independent means, given the observed standard deviations *(4 points)*
:::

####
::: {.callout}
# To do
With a power of 80%, $\alpha$=0.05, and measurements on 10 larvae at each site
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->


####
::: {.callout}
# To do
With a power of 80%, $\alpha$=0.05, and measurements on 200 larvae at each site
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->


####
::: {.callout}
# To do
With a power of 80%, $\alpha$=0.05, and measurements on 20 larvae at one site and 380 larvae at the second site
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->

####
::: {.callout}
# To do
Comment on how changing sample sizes affect `d`.
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->


### Differences in average mass

::: {.callout}
# To do
Calculate the difference in average mass, in $\mu$g, that can be detected given your estimate of the minimum detectable effect size (`d`) at 2.4.1-2.4.3 *(4 points)*
:::

<!--#################################################################-->
```{r}
#write your code here
```

Write your explanation

<!--#################################################################-->

