---
title: Bio4158/Bio5158 - Report 5
subtitle: Exercise
author: "Group 22 - Nalan Kurnaz"
date: today
format:
  html:
    embed-resources: true
    toc: true
    code-fold: show
    code-tools: true
---


# Background

Evason et al. (2005) tested the influence of an anticonvulsant medication on the lifespan of the rock star of all nematodes: *Caenorhabditis elegans*. The study **compared the lifespan of nematodes** under **three treatment levels** with a compound. Worms were either exposed to **water-only (the control)** or to the compound **at the larval stage only**, **at the adult stage**, or **at both larval and adult stages**.

# The data

The file `nematodes.csv` contains the results of this experiment. Each line corresponds to one worm. The two columns are the treatment level and the lifespan (in days). 

# The task

1. Analyze the data and determine if there is any evidence that the compound alters lifespan of the worms.
    a. The biological hypothesis and its predictions _(3 points)_
    b. The statistical null hypothesis for the omnibus test. _(2 points)_
    c. A relevant plot of the data separately by treatment level _(1 point)_
    d. Written description of the model to be fitted and provide a justification for the choice of analysis _(2 points)_
    e. Fit the model and provide evidence (plot, test, or statement) that you examined the statistical assumptions underlying a standard parametric analysis and a statement as to what you conclude. _(4 points)_
    f. The statistical conclusion (i.e. inference) you make from your analysis (i.e. do you accept or reject the null hypothesis?). Include follow-up tests, if appropriate, and explain what you conclude from them. _(4 points)_
    g. A statement about your confidence in the statistical conclusion (for example with respect to the violation of certain assumptions, to potential biases created by missing data, to suboptimal experimental design, and/or to power issues). _(1 point)_
    h. The ‘biological’ conclusion, meaning a qualified statement about worm lifespan, including an estimate of raw effect size, its precision, and whether an effect of this magnitude seems biologically relevant. _(2 points)_. 
2. Would your answers to 1f. change if you were later told that the authors had tested 18 other compounds? Explain. _(1 point)_

# Reference
Evason, K., C. Huang, I. Yamben, D. F. Covey, and K. Kornfeld. 2005. Anticonvulsant Medications Extend Worm Life-Span. Science 307: 258-262.


-------------------------------------

# Report

## 1. Analyze the data and determine if there is any evidence that the compound alters lifespan of the worms.
### a. The biological hypothesis and its predictions _(3 points)_

```{r}
#write your code here
```

Write your interpretation

### b. The statistical null hypothesis for the omnibus test. _(2 points)_

```{r}
#write your code here
```

Write your interpretation

### c. A relevant plot of the data separately by treatment level _(1 point)_

```{r}
#write your code here
```

Write your interpretation

### d. Written description of the model to be fitted and provide a justification for the choice of analysis _(2 points)_

```{r}
#write your code here
```

Write your interpretation

### e. Fit the model and provide evidence (plot, test, or statement) that you examined the statistical assumptions underlying a standard parametric analysis and a statement as to what you conclude. _(4 points)_

```{r}
#write your code here
```

Write your interpretation

### f. The statistical conclusion (i.e. inference) you make from your analysis (i.e. do you accept or reject the null hypothesis?). Include follow-up tests, if appropriate, and explain what you conclude from them. (4 points)

```{r}
#write your code here
```

Write your interpretation

### g. A statement about your confidence in the statistical conclusion (for example with respect to the violation of certain assumptions, to potential biases created by missing data, to suboptimal experimental design, and/or to power issues). (1 point)

```{r}
#write your code here
```

Write your interpretation

### h. The ‘biological’ conclusion, meaning a qualified statement about worm lifespan, including an estimate of raw effect size, its precision, and whether an effect of this magnitude seems biologically relevant. (2 points). 

```{r}
#write your code here
```

Write your interpretation

## 2. Would your answers to 1f. change if you were later told that the authors had tested 18 other compounds? Explain. (1 point)

```{r}
#write your code here
```

Write your interpretation

